---
title: "Getting Started with NASA's CMR-STAC API in R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

**This tutorial demonstrates how to interact with CMR-STAC in R.**

This tutorial will teach you how to navigate and explore NASA's Common Metadata Repository (CMR) SpatioTemporal Asset Catalog (STAC) to find and learn about the datasets available through NASA's different cloud archives. We'll demonstrate by using it to search for [ASTER Global DEM data](https://www.earthdata.nasa.gov/data/catalog/lpcloud-astgtm-003) available in the LP DAAC's Cumulus cloud archive.

------------------------------------------------------------------------

### Topics Covered in this Tutorial

1.  **Introduction to STAC and the CMR-STAC API**\
    1a. What is STAC?\
    1b. What is the CMR-STAC API?\
2.  **Get started & searching with CMR-STAC**\
    2a. CMR-STAC API\
    2b. STAC Collection\
    2c. STAC Item\
    2d. Assets

------------------------------------------------------------------------

### Required packages

-   **Required packages:**

    -   `dplyr`: manipulate data frames of stac search results
    -   `DT`: make returned information more readable
    -   `rstac`: interact with STAC catalogs
    -   `devtools`: install development version of packages
    -   `terra`: open DEM data once it has been discovered via STAC search

Run the cells below to install (or update) the necessary packages, and then load all of the required packages.

```{r}
#| eval: false

install.packages(c('dplyr','DT', 'rstac', 'devtools', 'terra'))
```

```{r}
library(dplyr)
library(DT)
library(rstac)
```

For listing the various STAC catalogs associated with NASA CMR, we'll install the development version of the `earthdatalogin` package in the following cell. The development version contains functions that will be utilized in section 2a.

```{r}
#| eval: false

devtools::install_github("boettiger-lab/earthdatalogin")
library("earthdatalogin")
```

------------------------------------------------------------------------

# 1. Introduction to STAC and the CMR-STAC API

## 1a. What is STAC?

STAC is short for [Spatiotemporal Asset Catalog](http://stacspec.org/), a series of specifications that standardize indexing and discovery of `spatiotemporal assets` (files containing information about the Earth across space and time).

There are four specifications that work both independently and together:

1)  [STAC Catalog](https://github.com/radiantearth/stac-spec/blob/master/catalog-spec/catalog-spec.md)
2)  [STAC Collection](https://github.com/radiantearth/stac-spec/blob/master/collection-spec/collection-spec.md)
3)  [STAC Item](https://github.com/radiantearth/stac-spec/blob/master/item-spec/item-spec.md)\
4)  [STAC API](https://github.com/radiantearth/stac-api-spec) specification builds on top of the three core specifications mentioned above. All these specifications are intended to be used together, yet are designed in a way that each piece is small, self-contained, and reusable in other contexts.

------------------------------------------------------------------------

## 1b. What is the CMR-STAC API?

The Common Metadata Repository (CMR) is a metadata system that catalogs Earth Science data and associated metadata records. NASA's CMR-STAC Application Programming Interface (API) is a translation API for STAC users who want to access and search through CMR's vast metadata holdings using STAC keywords.

------------------------------------------------------------------------

# 2. Get started with CMR-STAC

## 2a. CMR-STAC API

The CMR-STAC API contains endpoints that enable the querying of STAC items.

Here, we will use the function `list_nasa_stacs`, which connects to the CMR-STAC landing page (https://cmr.earthdata.nasa.gov/cloudstac/) for cloud datasets. The landing page contains all the available cloud data providers and their STAC endpoint. 

```{r}
cmr_cat_links <- earthdatalogin::list_nasa_stacs() |>
  jsonlite::toJSON(auto_unbox = TRUE) |> 
  jsonlite::fromJSON() |> 
  as.data.frame()
DT::datatable(cmr_cat_links)
```

The data frame above shows all the data providers with their associated STAC catalog endpoints. You will notice above that the CMR-STAC API contains many different endpoints--not just from NASA LP DAAC, but also contains endpoints for other NASA ESDIS DAACs. Use the `title` field to identify the data provider you are interested in. The data product used in this tutorial is hosted in the LP DAAC Cumulus Cloud space (LPCLOUD).

Let's get the associated endpoint for LPCLOUD.

```{r}
provider <- 'LPCLOUD'
lpcloud_cat_link <- earthdatalogin::get_nasa_stac_url(provider)
lpcloud_cat_link
```

------------------------------------------------------------------------

## 2b. STAC Collection

STAC Collection is an extension of STAC Catalog containing additional information that describes the STAC Items in that Collection.

Let's begin using the `rstac` package. `rstac` is a library that allows users to interact with STAC Catalogs and their associated data. Using the LPCLOUD link, we'll use `rstac` to query the LPCLOUD catalog and retrieve the content describing associated collections. Important information such as data collection ID and the title are provided here.

```{r}
lpcloud_collections <- stac(lpcloud_cat_link) |>
    collections() |>
    get_request()
print(lpcloud_collections)

for (i in lpcloud_collections) {
  collection_info <- lapply(lpcloud_collections$collections, function(x) {
      list(id = x$id, title = x$title)
  }) |>
  jsonlite::toJSON(auto_unbox = TRUE) |> 
  jsonlite::fromJSON() |> 
  as.data.frame()
  }

DT::datatable(collection_info)
```

In CMR, the Collection ID is used to query by a specific product, so be sure to save the ID for a collection you are interested in. For instance, the Collection ID for ASTER Global Digital Elevation Model V003 is `ASTGTM_003`. Note that the "id" shortname is in the format: productshortname_VVV (where VVV = product version).

Here, we use the short name `ASTGTM_003` to query the STAC Collection. If you are interested in querying a different LPCLOUD product, swap out the shortname to assign to the `collection` variable below.

Users can also define other paramaters such as a temporal and spatial extent. Notice the `limit` parameter in the `body` object. This parameter allows us to adjust the number of records returned during a request (default = 10).

```{r}
# define search parameters
collection <- 'ASTGTM_003'    
datetime <- '2000-01-01T00:00:00Z/2001-01-31T23:59:59Z'   #YYYY-MM-DDTHH:MM:SSZ/YYYY-MM-DDTHH:MM:SSZ
bbox <- c(-122.0622682571411,39.897234301806,-122.04918980598451,39.91309383703065) # LL and UR Coordinates

# search
collection_search <- stac(
    lpcloud_cat_link
)|>
  stac_search(
    limit = 100,
    collections = collection,
    bbox = bbox,
    datetime = datetime
  ) |>
  get_request()

print(collection_search)
```

We can see that the output includes the number of items that fall within the search criteria, the assets, and fields.

------------------------------------------------------------------------

## 2c. STAC Item

STAC Items represent data and metadata assets that are spatiotemporally coincident. Using our STAC Query, let's get the first item from our STAC collection search in the above cell.

```{r}
first_item <- collection_search$features[[1]]
first_item
```

------------------------------------------------------------------------

## 2d. Assets

The STAC Item ID (CMR Granule ID) is the unique identifier assigned to each granule within a data collection. Within each STAC Item are assets, which include the downloadable and streamable URL to data files along with other asset objects. Below, the first Granule ID is used to get associated files, which can be found in the 'href' field.

```{r}
item_assets <- first_item$assets

for (asset_name in names(item_assets)) {
    cat("Asset:", asset_name, "\n")
    cat("URL:", item_assets[[asset_name]]$href, "\n\n")
}
```

------------------------------------------------------------------------

The CMR-STAC Search endpoint allows user to quickly search for STAC Items that meet their specific spatial, temporal, and data product requirements. Now that you learned how to navigate and explore the CMR-STAC catalog, check out the [`HLS_tutorial`](https://git.earthdata.nasa.gov/projects/LPDUR/repos/hls_tutorial_r/raw/HLS_Tutorial.html?at=refs%2Fheads%2Fmain) to learn how to interact with HLS data specifically.

------------------------------------------------------------------------

### Contact Information
This tutorial was updated in July 2025 by the NASA IMPACT MAAP Team.

**Original material written by Mahsa Jami^1^ and Aaron Friesz^1^**\
Contact: [LPDAAC\@usgs.gov](mailto:LPDAAC@usgs.gov){.email}\
Voice: +1-866-573-3222\
Organization: Land Processes Distributed Active Archive Center (LP DAAC)\
Website: <https://lpdaac.usgs.gov/> \
Date last modified: 6/3/2025

^1^ KBR, Inc., contractor to the U.S. Geological Survey, Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, USA. Work performed under USGS contract G0121D0001 for LP DAAC^2^.

^2^ LP DAAC Work performed under NASA contract NNG14HH33I.
