---
title: Example Child Document
---

This chapter demonstrates how you can include a .qmd document inside another one. You can see that the following looks like a native part of this document, although it is actually a "child" document that is rendered inside this parent document. (The following is a test .qmd file. I copied it as a .ipynb and converted it to a .qmd file from <https://raw.githubusercontent.com/jupyter/notebook/master/docs/source/examples/Notebook/What%20is%20the%20Jupyter%20Notebook.ipynb>.) We can do the same with .ipynb files from NASA DAACs once we sort out the python environments. 

```{r, child="_python-example.qmd"}
```


## Our goal: PODAAC's notebook

This notebook demonstrates basic harmony subsetting from PO.DAAC: [**Harmony EOSS L2SS API Tutorial**](https://github.com/podaac/tutorials/blob/master/notebooks/harmony%20subsetting/Harmony%20L2%20Subsetter.ipynb).


Maybe this will be helpful for our python environments...

```{r setup, eval=FALSE}
#| echo: false

## install python environment: 
renv::restore(here("transformations"))

```

```{bash conda-env, eval=FALSE}
#| echo: false

## following https://quarto.org/docs/getting-started/quarto-projects.html

## create a new Python 3 virtual environment in the directory env:
#conda create --prefix env python 

# To use the environment you need to activate it.  
conda activate ./env 

## restore
conda create --prefix env -u -f environment.yml

```



