---
title: "How do I find data using R?"
execute:
  eval: false
---

Here are our recommended approaches for finding data with R, from the command line or a notebook.

## Authentication for NASA Earthdata 

An Earthdata Login account is required to access data from the NASA Earthdata system. Please visit <https://urs.earthdata.nasa.gov> to register and manage your Earthdata Login account. This account is free to create and only takes a moment to set up.

The first step is to get the correct authentication that will allow us to get cloud-hosted data from NASA. This is all done through Earthdata Login. We can use the `earthdatalogin` R package 
to help us manage this.

The easiest and most portable method of access is using the netrc basic
authentication protocol for HTTP. Call `edl_netrc()` to set this up
given your username and password (passed as optional arguments or read
from the environmental variables. If neither provides credentials,
`earthdatalogin` will provide it’s own credentials, but you may
experience rate limits more readily).

Once you have created an account on Earthdata, the easiest way to store your
credentials is by setting them as environment variables in your `~/.Renviron` file. 
The usethis package has a handy function to find and open this file for editing:

```r
usethis::edit_r_environ()
```

Then add the following lines to the file, save the file, and restart R

```
EARTHDATA_USER="your_user_name"
EARTHDATA_PASSWORD="your_password"
```

Now, when you call `edl_netrc()`, it will consult these environment variables
to create a netrc file that will be used to authenticate with the NASA Earthdata
servers.


``` r
edl_netrc()
```

If `edl_netrc()` has been called, then most existing spatial data
packages in R can then seamlessly access NASA Earthdata over HTTP links.

## Searching for data


The NASA cloud data is searchable programmatically via two methods - NASA's own search service, and the NASA Spatio-Temporal Asset Catalog (STAC) API. To find data in R, we'll mainly rely on the [rstac](https://brazil-data-cube.github.io/rstac/) package. This enables us interact with the NASA [STAC](https://stacspec.org/en) API to search for our data, and at the same time learn about STAC more generally. This will be useful as it is a common standard for distributing spatial data.

We will also search for data using the [NASA Earthdata search API](https://cmr.earthdata.nasa.gov/search/site/docs/search/api.html), which is the service that powers the [Earthdata web search portal](https://search.earthdata.nasa.gov/search?fst0=atmosphere).

For both of these services, the [earthdatalogin](https://boettiger-lab.github.io/earthdatalogin/) package will help us get set up and provide a few handy functions
to smooth some rough edges.

## rstac

All of the NASA STAC catalogues can be viewd here: https://radiantearth.github.io/stac-browser/#/external/cmr.earthdata.nasa.gov/stac/.

We will use the rstac package to first browse the collections in the PO DAAC 
catalogue (POCLOUD):

```{r}
## In R
## load R libraries
# install.packages("pak") 
# pak::pak(c("tidyverse", "rstac", "boettiger-lab/earthdatalogin"))

library(rstac)
library(earthdatalogin)

po_collections <- stac("https://cmr.earthdata.nasa.gov/stac/POCLOUD/") |>
  collections() |>
  get_request()

po_collections
```

This only gives us the first 10 collections in the catalogue; to get the rest
we can use the `collections_fetch()` function:

```{r}
all_po_collections <- collections_fetch(po_collections)
all_po_collections

length(all_po_collections$collections)

head(all_po_collections$collections)
```

Once we have searched through the collections, we can choose the one we are 
interested in and query it to get the items (granules) we are interested in:

```{r}
items <- stac("https://cmr.earthdata.nasa.gov/stac/POCLOUD") |> 
  stac_search(collections = "MUR-JPL-L4-GLOB-v4.1",
              datetime = paste(start,end, sep = "/")) |>
  post_request() |>
  items_fetch()

items
```




```{r}
granules <- edl_search(
  doi = URLencode("10.5067/SLREF-CDRV3", reserved = TRUE)
)

## Granules found: 72

## exploring
granules # this is the result of the get request. 

class(granules) # "list"
## granules <- reticulate::py_to_r(granules) # Object to convert is not a Python object

```
