---
title: "Find Data: Programmatic Search"
---

## Introduction

We can find data programmatically using the following code.

## Code

Here are our recommended approaches for finding data with code.

::: {.panel-tabset group="language"}
## Python

In Python we can use the [`earthdata`](https://earthdata.readthedocs.io/en/latest/) library (to be renamed `earthaccess` very soon!)

To install the package we'll run this code from the command line. Note: you can run shell code directly from a Jupyter Notebook cell by adding a `!`: `!conda install`.

```{bash}
#| eval: false
# Install earthdata
conda install -c conda-forge earthdata
```

This example searches for data from the LPDAAC with a spatial bounding box. 

```{python}
#| eval: false

## Import earthdata
from earthdata import DataGranules

# Then we build a Query with spatiotemporal parameters
GranuleQuery = DataGranules().concept_id("C1575731655-LPDAAC_ECS").bounding_box(-134.7,58.9,-133.9,59.2)

# We get the metadata records from CMR
granules = GranuleQuery.get()
```




## R

R code coming soon!

```{r}
# Coming soon!
```

## Matlab

Matlab code coming soon!

```{r}
# Coming soon!
```

## Command Line

With `wget` and `curl`:

```{r}
# Coming soon!
```


:::
